---
title: "Injury Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Injury Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/mrsimple07/injury-prediction-dataset\>*

### Reference:

*\<mrsimple07. (n.d.). Injury Prediction Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/mrsimple07/injury-prediction-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Loading Dataset}
# Load the dataset
InjuryData <- read.csv("injury_data.csv", colClasses = c(
  Player_Age = "numeric",
  Player_Weight = "numeric",
  Player_Height = "numeric",
  Previous_Injuries = "factor",
  Training_Intensity = "numeric",
  Recovery_Time = "numeric",
  Likelihood_of_Injury = "factor"  # Assuming this is the target variable (categorical)
))
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency of categorical variable
freq_table <- table(InjuryData$Likelihood_of_Injury)
print("Frequency Table:")
print(freq_table)

# Proportions of categories
prop_table <- prop.table(freq_table)
print("Proportions Table:")
print(prop_table)

# Summary of numeric variables (count of non-missing values)
print("Summary of Player_Age and Training_Intensity:")
summary(InjuryData$Player_Age)
summary(InjuryData$Training_Intensity)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean
mean_age <- mean(InjuryData$Player_Age, na.rm = TRUE)
mean_weight <- mean(InjuryData$Player_Weight, na.rm = TRUE)
print("Mean of Player_Age and Player_Weight:")
print(mean_age)
print(mean_weight)

# Median
median_age <- median(InjuryData$Player_Age, na.rm = TRUE)
median_weight <- median(InjuryData$Player_Weight, na.rm = TRUE)
print("Median of Player_Age and Player_Weight:")
print(median_age)
print(median_weight)

# Mode (custom function, as R does not have a built-in mode function)
get_mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}
mode_injuries <- get_mode(InjuryData$Previous_Injuries)
print("Mode of Previous_Injuries:")
print(mode_injuries)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Range
range_age <- range(InjuryData$Player_Age, na.rm = TRUE)
range_weight <- range(InjuryData$Player_Weight, na.rm = TRUE)
print("Range of Player_Age and Player_Weight:")
print(range_age)
print(range_weight)

# Variance
var_age <- var(InjuryData$Player_Age, na.rm = TRUE)
var_weight <- var(InjuryData$Player_Weight, na.rm = TRUE)
print("Variance of Player_Age and Player_Weight:")
print(var_age)
print(var_weight)

# Standard Deviation
sd_age <- sd(InjuryData$Player_Age, na.rm = TRUE)
sd_weight <- sd(InjuryData$Player_Weight, na.rm = TRUE)
print("Standard Deviation of Player_Age and Player_Weight:")
print(sd_age)
print(sd_weight)

# Skewness and Kurtosis (requires 'e1071' package)
library(e1071)

skew_age <- skewness(InjuryData$Player_Age, na.rm = TRUE)
kurt_age <- kurtosis(InjuryData$Player_Age, na.rm = TRUE)
print("Skewness and Kurtosis of Player_Age:")
print(skew_age)
print(kurt_age)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between two numeric variables
cor_age_weight <- cor(InjuryData$Player_Age, InjuryData$Player_Weight, use = "complete.obs")
print("Correlation between Player_Age and Player_Weight:")
print(cor_age_weight)

# Covariance between two numeric variables
cov_age_weight <- cov(InjuryData$Player_Age, InjuryData$Player_Weight, use = "complete.obs")
print("Covariance between Player_Age and Player_Weight:")
print(cov_age_weight)

# Scatterplot to visualize relationship
plot(InjuryData$Player_Age, InjuryData$Player_Weight, 
     xlab = "Player Age", ylab = "Player Weight", main = "Age vs Weight")
```

## ANOVA
```{r ANOVA}
# Perform ANOVA: Testing if Player_Age differs by Likelihood_of_Injury
anova_age <- aov(Player_Age ~ Likelihood_of_Injury, data = InjuryData)

# Print summary of ANOVA for Player_Age
print("ANOVA for Player_Age by Likelihood_of_Injury:")
summary(anova_age)

# Perform ANOVA: Testing if Player_Weight differs by Likelihood_of_Injury
anova_weight <- aov(Player_Weight ~ Likelihood_of_Injury, data = InjuryData)

# Print summary of ANOVA for Player_Weight
print("ANOVA for Player_Weight by Likelihood_of_Injury:")
summary(anova_weight)
```

## Plots
```{r Plots}
# Boxplot for Player_Age by Likelihood_of_Injury
boxplot(Player_Age ~ Likelihood_of_Injury, data = InjuryData,
        main = "Boxplot of Player_Age by Likelihood_of_Injury",
        xlab = "Likelihood of Injury", ylab = "Player Age")

# Boxplot for Player_Weight by Likelihood_of_Injury
boxplot(Player_Weight ~ Likelihood_of_Injury, data = InjuryData,
        main = "Boxplot of Player_Weight by Likelihood_of_Injury",
        xlab = "Likelihood of Injury", ylab = "Player Weight")

# Univariate Plots
# Histogram for Player_Age
hist(InjuryData$Player_Age, 
     main = "Histogram of Player_Age", 
     xlab = "Player Age", 
     col = "lightblue", 
     border = "black", 
     breaks = 10)

# Histogram for Player_Weight
hist(InjuryData$Player_Weight, 
     main = "Histogram of Player_Weight", 
     xlab = "Player Weight", 
     col = "lightgreen", 
     border = "black", 
     breaks = 10)

# Boxplot for Player_Age
boxplot(InjuryData$Player_Age, 
        main = "Boxplot of Player_Age", 
        ylab = "Player Age", 
        col = "lightblue", 
        border = "black")

# Boxplot for Player_Weight
boxplot(InjuryData$Player_Weight, 
        main = "Boxplot of Player_Weight", 
        ylab = "Player Weight", 
        col = "lightgreen", 
        border = "black")

# Boxplot for Player_Age by Likelihood_of_Injury
boxplot(Player_Age ~ Likelihood_of_Injury, 
        data = InjuryData, 
        main = "Boxplot of Player_Age by Likelihood_of_Injury", 
        xlab = "Likelihood of Injury", 
        ylab = "Player Age", 
        col = c("lightblue", "lightgreen"))

# Multivariate Plots
# Scatterplot between Player_Age and Player_Weight
plot(InjuryData$Player_Age, InjuryData$Player_Weight, 
     main = "Scatterplot of Player_Age vs Player_Weight", 
     xlab = "Player Age", 
     ylab = "Player Weight", 
     col = "blue", 
     pch = 16)

# Pairwise scatterplot matrix for numerical variables
pairs(InjuryData[, c("Player_Age", "Player_Weight", "Training_Intensity", "Recovery_Time")],
      main = "Pairs Plot of Numerical Variables")

# Facet grid plot for numerical variables by Likelihood_of_Injury using 'ggplot2'
library(ggplot2)

# Scatter plot of Player_Age vs Player_Weight, faceted by Likelihood_of_Injury
ggplot(InjuryData, aes(x = Player_Age, y = Player_Weight, color = Likelihood_of_Injury)) +
  geom_point() +
  facet_wrap(~Likelihood_of_Injury) +
  labs(title = "Player_Age vs Player_Weight by Likelihood_of_Injury",
       x = "Player Age", y = "Player Weight") +
  theme_minimal()

# Boxplot of Player_Age by Likelihood_of_Injury using 'ggplot2'
ggplot(InjuryData, aes(x = Likelihood_of_Injury, y = Player_Age, fill = Likelihood_of_Injury)) +
  geom_boxplot() +
  labs(title = "Boxplot of Player_Age by Likelihood_of_Injury",
       x = "Likelihood of Injury", y = "Player Age") +
  theme_minimal()

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check if there are any missing values in the dataset
# This will return TRUE for any column that has missing values
missing_values <- sapply(InjuryData, function(x) sum(is.na(x)))

# Display the number of missing values per column
print("Missing values per column:")
print(missing_values)

# Check if there are any missing values in the entire dataset
total_missing <- sum(is.na(InjuryData))
cat("Total missing values in the dataset:", total_missing, "\n")

# Check if any column has missing values
any_missing <- any(missing_values > 0)
cat("Does the dataset have missing values?", ifelse(any_missing, "Yes", "No"), "\n")

# Visualize missing data (optional, using the 'VIM' package)
library(VIM)
aggr(InjuryData, col = c("navyblue", "yellow"), numbers = TRUE, sortVars = TRUE, labels = names(InjuryData), cex.axis = 0.7, gap = 3, ylab = c("Missing data", "Pattern"))

```

# Training Model
## Data Splitting
```{r Data Splitting}
library(caret)

# Set seed for reproducibility
set.seed(123)

# 1. Data Splitting (80% Training, 20% Testing)
train_index <- sample(1:nrow(InjuryData), 0.8 * nrow(InjuryData))
train_data <- InjuryData[train_index, ]
test_data <- InjuryData[-train_index, ]
cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")

```

## Bootstrapping
```{r Bootstrapping}
# 2. Bootstrapping (Sampling with replacement)
n_bootstrap <- 1000
bootstrap_samples <- matrix(NA, nrow = n_bootstrap, ncol = nrow(InjuryData))
for(i in 1:n_bootstrap) {
  bootstrap_samples[i, ] <- sample(1:nrow(InjuryData), nrow(InjuryData), replace = TRUE)
}
head(bootstrap_samples)
```

## Cross-validation
```{r Cross-validation}
# 3. Cross-Validation (Caret package)
library(caret)

# a. Basic k-Fold Cross-Validation (10-fold)
train_control <- trainControl(method = "cv", number = 10)
model_cv <- train(Likelihood_of_Injury ~ ., data = train_data, method = "glm", family = "binomial", trControl = train_control)
print(model_cv)

# b. Repeated k-Fold Cross-Validation (10-fold, repeating 3 times)
train_control_repeated <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
model_repeated_cv <- train(Likelihood_of_Injury ~ ., data = train_data, method = "glm", family = "binomial", trControl = train_control_repeated)
print(model_repeated_cv)

# c. Leave-One-Out Cross-Validation (LOOCV)
train_control_loocv <- trainControl(method = "LOOCV")
model_loocv <- train(Likelihood_of_Injury ~ ., data = train_data, method = "glm", family = "binomial", trControl = train_control_loocv)
print(model_loocv)

```

## Training Different Models
```{r Training Different Models}
# 4. Model Training with Decision Tree (rpart)
model_dt <- train(Likelihood_of_Injury ~ ., data = train_data, method = "rpart", trControl = train_control)
print(model_dt)

# 5. Model Training with Random Forest (rf)
model_rf <- train(Likelihood_of_Injury ~ ., data = train_data, method = "rf", trControl = train_control)
print(model_rf)
```

## Performance Comparison with resamples
```{r}
# 3. Model Performance Comparison using resamples
models_list <- resamples(list(GLM = model_cv, DecisionTree = model_dt, RandomForest = model_rf))

# Summarize the resamples for comparison
summary(models_list)

# Boxplot for visual comparison
bwplot(models_list)
```

## Saving Model
```{r Saving Model}
# Save the Random Forest model to an RDS file
saveRDS(model_rf, "./models/saved_rf_model.rds")

# Load the saved Random Forest model
loaded_rf_model <- readRDS("./models/saved_rf_model.rds")

# Example new data for prediction
new_data <- data.frame(
  Player_Age = 24,
  Player_Weight = 66.25,
  Player_Height = 175.73,
  Previous_Injuries = factor(1, levels = c(0, 1)),
  Training_Intensity = 5,
  Recovery_Time = 0.46,
  Likelihood_of_Injury = factor(0, levels = c(0, 1))
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print the predictions
print(predictions_loaded_model)

```

